---
execute: 
  echo: false
  message: false
  warning: false
---
## Inroduction

The World Happiness Report reviews the state of happiness in the world today and shows how the "science of happiness" explains personal and national variations in happiness.
The scores are based on peoples assessments of their lives in each country.
Individuals are asked the Cantril Ladder question:
*Imagining a ladder with 10 steps, 10 being the best possible life, on which step do you see yourself?*

This project looks at the latitude of each country to try and determine if there is a relationship between
the latitude of a country and it's Ladder score.

Data Sources:

* The World Happiness Report: <https://worldhappiness.report>
* Global country boundaries (ADMIN0) <https://www.naturalearthdata.com/downloads/>
 

```{r}
#| label: setup

library(tidyverse)
library(sf)
library(sp)
library(readxl)

```

```{r}
#| label: dataloading

# load country layer
countries <- st_read("natural_earth_vector.gpkg",
                     layer = "ne_10m_admin_0_countries",
                     quiet = TRUE)
country_geom <- countries |>
  select(ADMIN, geom)

# load World Happiness indicators
whr <- read_excel("DataForFigure2.1+with+sub+bars+2024.xls")
whr_subset <- whr |>
  rename(country_name = `Country name`,
         score = `Ladder score`) |>
  select(country_name, score)

# load table matching country names between above 2 datasets
name_match <- read_csv("country_name_match.csv")

```


```{r}
#| label: datamunging
# function to calculate country mid-latitude
get_mid_latitude <- function(geometry) {
  coords <- st_coordinates(geometry)
  min_lat <- min(coords[, 2])
  max_lat <- max(coords[, 2])
  mid_lat <- ((max_lat - min_lat) / 2) + min_lat
  return(mid_lat)
}

# merge Wold Happiness indcators with geometry from country layer
whr_withgeom <- country_geom |>
  merge(name_match, by = "ADMIN", all.x = TRUE) |>
  merge(whr_subset, by = "country_name", all.x = TRUE)

# calculate mid-latitude per country
whr_withgeom$mid_latitude <- whr_withgeom |>
  apply(1, function(row) get_mid_latitude(row$geometry))

```

## Mapping Happiness Scores
Simply mapping It is difficult to get a clear picture of global happiness from simply mapping happiness scores.

```{r}
#| label: worldmap_scores

# Plot the choropleth map

latitude_labels <- seq(-90, 90, by = 30)
longitude_labels <- seq(-180, 180, by = 45)
ggplot() +
  geom_sf(data = whr_withgeom, aes(fill = score)) +
    scale_fill_gradient(low = "lightblue",
                      high = "darkblue",
                      na.value = "grey96",
                      name = "Score") +
    labs(title = "Happiness Score by Country",
         subtitle = "No data for grey areas",
         caption = "Source: World Happiness Report, 2024") +
    theme_minimal() 

```

## Graphing Happiness Scores with Mid-latitude

```{r}
#| label: latscore_graph
whr_withgeom |>
  ggplot(aes(x = score, y = mid_latitude)) +
    geom_point() +
    geom_smooth(method = "lm", formula = "y ~ x") +
    scale_y_continuous(name = "Mid latitude",
                       breaks = c(-60, -30, 0, 30, 60),
                       limits = c(-80, 80)) +
    scale_x_continuous(name = "World Happiness Score",
                       breaks = seq(1,8)) +
    labs(title = "World Happiness Score by Country Latitude",
         subtitle = "Higher Latitiudes have slightly higher happiness scores",
         caption = "Source: https://worldhappiness.report, 2024")

```

